{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}

### 2016-08-22 ident_manu.Rnw
explanation of parameter removal

This study evaluates a recently developed hydrodynamic and ecological model that describes horizontal and vertical transport and mixing of state variables relevant for hypoxia in the northern \ac{gom}.  The \ac{cgem} includes elements from the Navy Coastal Ocean Model \citep{Martin00} that describe hydrodynamics on the \ac{lcs} and a biogeochemical model with multiple plankton groups, water-column metabolism, and sediment diagenesis \citep{Eldridge10}.  The hydrodynamic component of \ac{cgem} provides a spatially-explicit description of hypoxia using an orthogonal grid with an approximate horizontal resolution of 1.9 km$^2$ and twenty equally-spaced vertical sigma layers on the shelf (depth $\leq$ 100 m, with additional hybrid layers at deeper depths).  The biogeochemical component includes equations for 36 state variables including six phytoplankton groups (with nitrogen and phosophorus quotas for each), two zooplankton groups, nitrate, ammonium, phosphate, dissolved inorganic carbon, oxygen, silica, and multiple variables for dissolved and particulate organic matter from different sources.  Atmospheric and hydrological boundary conditions described in \citet{Hodur97} and \citet{Lehrter13} are also included in \ac{cgem}.

The core unit of \ac{cgem} is FishTank, a \ac{zerod} model that implements the biogeochemical equations in \citet{Eldridge10} and does not include any form of physical transport (i.e., advection, mixing, or surface flux).  Although FishTank was developed for specific application in \ac{cgem}, it can easily be applied to other hydrodynamic grids. Accordingly, the sensitivity and identifiability analysis described below are informative for both the \ac{lcs} gridded model as well as potential applications to different systems.  The FishTank model provides estimates for the 36 state variables described above using a \ac{zerod} parcel that is uniformly mixed as a closed system.  A set of initial conditions is provided as input to the model that was based on observations of relevant variables obtained from research cruises in April, June, and September 2006 \citet[Table 1 in][]{Murrell14}.

<<echo = FALSE>>=
# numer of parameters in each category
nparms <- table(parcats2(as_df = TRUE)$cat) %>% 
  as.list
@

Results from FishTank are based on time-dependent differential equations that describe energy flow between phytoplankton and zooplankton groups as affected by nutrient uptake rates, organic matter inputs and losses, inherent optical properties, sediment diagenesis, and temperature (\citealt{Penta08,Eldridge10}, see appendix in \citetalias{LehrterIR}).  A total of 108 equations are estimated at each time step to return a value for each of the 36 state variables described by the model.  In addition to the initial conditions, 250 parameter values for each of the equations is also supplied at model execution.  These parameters define relationships among fixed effects in the equations and represent ecological properties described by the model that influence hypoxia formation.  Values for each of the parameters were based on estimates from the literature, field or laboratory-based measurements, or expert knowledge in absence of the former.  As such, a sensitivity analysis of parameter values is warranted given that, for example, literature or field-based estimates may not apply under all scenarios or expert knowledge is not completely certain \citep{Refsgaard07}.  The sensitivity of \ac{do} to perturbations of all parameters for the 108 equations was estimated from January 1\textsuperscript{st} to December 31\textsuperscript{st}, 2006 by running FishTank at a timestep of five minutes.  For simplicity, the parameters were grouped into one of six categories based on their respective equations: optics ($n = $ \Sexpr{nparms$Optics} parameters), organic matter (\Sexpr{nparms$`Organic Matter`}), phytoplankton (\Sexpr{nparms$Phytoplankton}), temperature (\Sexpr{nparms$Temperature}), and zooplankton (\Sexpr{nparms$Zooplankton}).  A full description is available as an appendix in \citepalias{LehrterIR}.

@@

The analysis focused on sensitivity of \ac{do} in the \ac{zerod} FishTank model to identify parameters that may affect spatial and temporal variation of hypoxia in the larger model.  A local sensitivity analysis was performed for each of the 250 parameters using a simple perturbation approach to evaluate the change in \ac{do} from the original parameter values.  The analyses relied exlusively on concepts used in the FME package developed for the R statistical programming language \citep{Soetaert10}. Each parameter was perturbed by 50\% of its original value and the model was executed to obtain an estimate of the effect on \ac{do}.  For each perturbation, a sensitivity value $S$ was estimated for each time step $i$ given a set value for parameter $j$ as:

@@

\noindent where the estimate depended on the change in the predicted value for response variable $y$ divided by the change in the parameter $\Theta_j$ multiplied by the quotient of scaling factors $w$ for each.  The scaling factors, $w_{\Theta_j}$ for the parameter $\Theta_j$ and $w_{y_i}$ for response variable $y_i$, were set as the default value of the unperturbed parameter and the predicted value of $y_i$ after perturbation \citep{Soetaert10}.  The scaling ensures the estimates are unitless such that the relative magnitudes provide a comparison for model sensitivity to parameter changes that may vary in scale.  Estimates for $S_{ij}$ were summarized as $L1$ and $L2$ across the time series to obtain individual sensitivity values of \ac{do} in response to a change in parameter $j$:

@@

Initial analyses suggested that considerably limited subsets of parameters were identifiable of the 250 in the FishTank model.  Given this limitation, parameter selection must consider the competing objectives of increased precision with parameter inclusion and reduced identifability as it relates to optimization.  An additional challenge is the excessively high number of combinations of parameter sets, which complicates selection given sensitivity differences and desired ecological categories of each parameter.  For example, \cref{fig:combnex} provides a simple graphic of the unique number of combinations that are possible for different subsets of `complete' parameter sets of different sizes (i.e., based on $n$ choose $k$ combinations equal to $n!/\left(k!\left(n-k\right)!\right)$).  The number of unique combinations increases with the total parameters in the set and is also maximized for moderate selections (e.g., selecting half the total).  For example, over 10$^{14}$ combinations are possible by selecting 25 parameters from a set of 50.  Accordingly, parameter selection is complicated by differing sensitivity, identifiability, and the difficulty of choosing from many combinations.

@@

Local sensitivity analyses showed that \ac{do} was sensitive to perturbations in \Sexpr{nrow(sens_ests_cat)} of 250 parameters (\Sexpr{round(100 * nrow(sens_ests_cat)/250, 0)}\% of total) in FishTank (see \cref{fig:sensalltile} for other state variables). Within each parameter category, \ac{do} was sensitive to \Sexpr{bycat$Optics} parameters for optics (\Sexpr{lncats$Optics} of all optic parameters, \cref{tab:optsens}), \Sexpr{bycat$`Organic Matter`} for organic matter (\Sexpr{lncats$`Organic Matter`}, \cref{tab:omsens}), \Sexpr{bycat$Phytoplankton} for phytoplankton (\Sexpr{lncats$Phytoplankton}, \cref{tab:phytosens}), \Sexpr{bycat$Temperature} for temperature (\Sexpr{lncats$Temperature}, \cref{tab:tempsens}), and \Sexpr{bycat$Zooplankton} for zooplankton (\Sexpr{lncats$Zooplankton}, \cref{tab:zoopsens}). Although \ac{do} had the greatest sensitivity to parameters in the zooplankton category (as percentage of total), the relative effects varied. Among all parameters, sensitivity values ranged from $L1 = $ \Sexpr{sensrng$error[1]} for \Sexpr{sensrng$Parameter[1]} (\Sexpr{tolower(sensrng$Category[1])}) to \Sexpr{sensrng$error[2]} for \Sexpr{sensrng$Parameter[2]} (\Sexpr{tolower(sensrng$Category[2])}), whereas average sensitivity among all parameters was $L1 = $ \Sexpr{sensave}.  Within categories, sensitivity ranged from \Sexpr{sensrngcat$Optics$error[1]} (\Sexpr{sensrngcat$Optics$Parameter[1]}) to \Sexpr{sensrngcat$Optics$error[2]} (\Sexpr{sensrngcat$Optics$Parameter[2]}) for optics, \Sexpr{sensrngcat$`Organic Matter`$error[1]} (\Sexpr{sensrngcat$`Organic Matter`$Parameter[1]}) to \Sexpr{sensrngcat$`Organic Matter`$error[2]} (\Sexpr{sensrngcat$`Organic Matter`$Parameter[2]}) for organic matter, \Sexpr{sensrngcat$Phytoplankton$error[1]} (\Sexpr{sensrngcat$Phytoplankton$Parameter[1]}) to \Sexpr{sensrngcat$Phytoplankton$error[2]} (\Sexpr{sensrngcat$Phytoplankton$Parameter[2]}) for phytopankton, \Sexpr{sensrngcat$Temperature$error[1]} (\Sexpr{sensrngcat$Temperature$Parameter[1]}) to \Sexpr{sensrngcat$Temperature$error[2]} (\Sexpr{sensrngcat$Temperature$Parameter[2]}) for temperature, and \Sexpr{sensrngcat$Zooplankton$error[1]} (\Sexpr{sensrngcat$Zooplankton$Parameter[1]}) to \Sexpr{sensrngcat$Zooplankton$error[2]} (\Sexpr{sensrngcat$Zooplankton$Parameter[2]}) for zooplankton (\cref{fig:sensplo}, bottom).  Average sensitivity values in each category were $L1 = $ \Sexpr{sensave_cat$Optics} for optics, \Sexpr{sensave_cat$`Organic Matter`} for organic matter, \Sexpr{sensave_cat$Phytoplankton} for phytoplankton, \Sexpr{sensave_cat$Temperature} for temperature, and \Sexpr{sensave_cat$Zooplankton} for zooplankton.  Within the six phytoplankton groups, \ac{do} was sensitive to the same parameters within each group although the sensitivity magnitudes varied. Average sensitivity across parameters in each phytoplankton groups showed that \ac{do} was most sensitive to the first and third phytoplankton groups (average $L1 = $ \Sexpr{phytogrps[1, 2]}, \Sexpr{phytogrps[2, 2]}), whereas sensitivity to parameters in the remaining groups was much lower (all with average $L1 < 1$). Sensitivity of \ac{do} did not vary considerably between parameters in the two zooplankton groups (average $L1 = $ \Sexpr{zoopgrps[1, 2]}, \Sexpr{zoopgrps[2, 2]} for groups one and two).

>>>>

This study evaluated a recently developed hydrodynamic and ecological model that describes horizontal and vertical transport and mixing of state variables relevant for hypoxia in the northern \ac{gom}.  The \ac{cgem} includes elements from the Navy Coastal Ocean Model \citep{Martin00} that describes hydrodynamics on the \ac{lcs} and a biogeochemical model with multiple plankton groups, water-column metabolism, and sediment diagenesis \citep{Eldridge10}.  The hydrodynamic component of \ac{cgem} provides a spatially-explicit description of hypoxia using an orthogonal grid with an approximate horizontal resolution of 1.9 km$^2$ and twenty equally-spaced vertical sigma layers on the shelf (depth $\leq$ 100 m, with additional hybrid layers at deeper depths).  The biogeochemical component includes equations for 36 state variables including six phytoplankton groups (with nitrogen and phosophorus quotas for each), two zooplankton groups, nitrate, ammonium, phosphate, dissolved inorganic carbon, oxygen, silica, and multiple variables for dissolved and particulate organic matter from different sources.  Atmospheric and hydrological boundary conditions described in \citet{Hodur97} and \citet{Lehrter13} are also included in \ac{cgem}.

The core unit of \ac{cgem} is FishTank, a \ac{zerod} model that implements the biogeochemical equations in \citet{Eldridge10} and does not include any form of physical transport (i.e., advection, mixing, or surface flux).  Although FishTank was developed for specific application in \ac{cgem}, it can easily be applied to other hydrodynamic grids. Accordingly, the sensitivity and identifiability analysis described below are informative for both the \ac{lcs} gridded model as well as potential applications to different systems.  The FishTank model provides estimates for the 36 state variables described above using a \ac{zerod} parcel that is uniformly mixed as a closed system.  A set of initial conditions is provided as input to the model that was based on observations of relevant variables obtained from research cruises in April, June, and September 2006 \citep[Table 1 in][]{Murrell14}.

<<echo = FALSE>>=
# remove irrelevant parameters, from 'ignore/noparams.csv'
rmpars <- c('Kw_1', 'Kcdom_1', 'Kspm_1', 'Kchla_1', 'aw490_1', 'astarOMR_1', 'astarOMBC_1', 'KTg1(nospA+nospZ)_1', 'KTg1(nospA+nospZ)_2', 'KTg1(nospA+nospZ)_3', 'KTg1(nospA+nospZ)_4', 'KTg1(nospA+nospZ)_5', 'KTg1(nospA+nospZ)_6', 'KTg1(nospA+nospZ)_7', 'KTg1(nospA+nospZ)_8', 'KTg2(nospA+nospZ)_1', 'KTg2(nospA+nospZ)_2', 'KTg2(nospA+nospZ)_3', 'KTg2(nospA+nospZ)_4', 'KTg2(nospA+nospZ)_5', 'KTg2(nospA+nospZ)_6', 'KTg2(nospA+nospZ)_7', 'KTg2(nospA+nospZ)_8', 'Ea_R(nospA+nospZ)_1', 'Ea_R(nospA+nospZ)_2', 'Ea_R(nospA+nospZ)_3', 'Ea_R(nospA+nospZ)_4', 'Ea_R(nospA+nospZ)_5', 'Ea_R(nospA+nospZ)_6', 'Ea_R(nospA+nospZ)_7', 'Ea_R(nospA+nospZ)_8', 'beta_1', 'beta_2', 'beta_3', 'beta_4', 'beta_5', 'beta_6', 'QmaxN_1', 'QmaxN_2', 'QmaxN_3', 'QmaxN_4', 'QmaxN_5', 'QmaxN_6', 'QmaxP_1', 'QmaxP_2', 'QmaxP_3', 'QmaxP_4', 'QmaxP_5', 'QmaxP_6', 'nfQs_1', 'nfQs_2', 'nfQs_3', 'nfQs_4', 'nfQs_5', 'nfQs_6', 'KG1_R_1', 'KG2_R_1', 'KG1_BC_1', 'KG2_BC_1', 'pCO2_1', 'stoich_x1R_1', 'stoich_y1R_1', 'stoich_x2R_1', 'stoich_y2R_1', 'stoich_x1BC_1', 'stoich_y1BC_1', 'stoich_x2BC_1', 'stoich_y2BC_1', 'sink OM1_R_1', 'sink OM2_R_1', 'sink OM1_BC_1', 'sink OM2_BC_1', 'CF_SPM_1')

# number of parameters in each category
nparms <- parcats2(as_df = TRUE) %>% 
  filter(!shrt %in% rmpars) %>% 
  .$cats %>% 
  table %>% 
  as.list
@

Results from FishTank are based on time-dependent differential equations that describe energy flow between phytoplankton and zooplankton groups as affected by nutrient uptake rates, organic matter inputs and losses, inherent optical properties, sediment diagenesis, and temperature (\citealt{Penta08,Eldridge10}, see appendix in \citetalias{LehrterIR}).  A total of 108 equations are estimated at each time step to return a value for each of the 36 state variables described by the model.  In addition to the initial conditions, 251 parameter values for each of the equations are also supplied at model execution.  These parameters define relationships among fixed effects in the equations and represent ecological properties described by the model that influence hypoxia formation.  Values for each of the parameters were based on estimates from the literature, field or laboratory-based measurements, or expert knowledge in absence of the former.  As such, a sensitivity analysis of parameter values is warranted given that, for example, literature or field-based estimates may not apply under all scenarios or expert knowledge is not completely certain \citep{Refsgaard07}.  The sensitivity of \ac{do} to perturbations of all relevant parameters for the 108 equations was estimated from January 1\textsuperscript{st} to December 31\textsuperscript{st}, 2006 by running FishTank at a timestep of five minutes.  Initial screening removed 73 irrelevant parameters that were not relevant such that 178 were retained for the sensitivity analysis.  Irrelevant parameters were removed for several reasons, e.g., a parameter was only relevant for the hydrodynamic model, was considered a physical constant, or had no effect given initial conditions.  For simplicity, the remaining parameters were grouped into one of six categories based on their respective equations: optics ($n = $ \Sexpr{nparms$Optics} parameters), organic matter (\Sexpr{nparms$`Organic Matter`}), phytoplankton (\Sexpr{nparms$Phytoplankton}), temperature (\Sexpr{nparms$Temperature}), and zooplankton (\Sexpr{nparms$Zooplankton}).  A full description is available as an appendix in \citepalias{LehrterIR}.

@@

The analysis focused on sensitivity of \ac{do} in the \ac{zerod} FishTank model to identify parameters that may affect spatial and temporal variation of hypoxia in the larger model.  A local sensitivity analysis was performed for each of the 178 parameters using a simple perturbation approach to evaluate the change in \ac{do} from the original parameter values.  The analyses relied exlusively on concepts used in the FME package developed for the R statistical programming language \citep{Soetaert10}. Each parameter was perturbed by 50\% of its original value and the model was executed to obtain an estimate of the effect on \ac{do}.  For each perturbation, a sensitivity value $S$ was estimated for each time step $i$ given a set value for parameter $j$ as:

@@

\noindent where the estimate depended on the change in the predicted value for response variable $y$ divided by the change in the parameter $\Theta_j$ multiplied by the quotient of scaling factors $w$ for each.  The scaling factors, $w_{\Theta_j}$ for the parameter $\Theta_j$ and $w_{y_i}$ for response variable $y_i$, were set as the default value of the unperturbed parameter and the predicted value of $y_i$ after perturbation \citep{Soetaert10}.  The scaling ensures the estimates are unitless such that the relative magnitudes provide a comparison for model sensitivity to parameter changes that may vary in scale.  Estimates for $S_{ij}$ were summarized as $L1$ and $L2$ across the time series to obtain individual sensitivity values of \ac{do} in response to a change in parame ter $j$:

@@

Initial analyses suggested that considerably limited subsets of parameters were identifiable of the 178 evaluated for the FishTank model.  Given this limitation, parameter selection must consider the competing objectives of increased precision with parameter inclusion and reduced identifability as it relates to optimization.  An additional challenge is the excessively high number of combinations of parameter sets, which complicates selection given sensitivity differences and desired ecological categories of each parameter.  For example, \cref{fig:combnex} provides a simple graphic of the unique number of combinations that are possible for different subsets of `complete' parameter sets of different sizes (i.e., based on $n$ choose $k$ combinations equal to $n!/\left(k!\left(n-k\right)!\right)$).  The number of unique combinations increases with the total parameters in the set and is also maximized for moderate selections (e.g., selecting half the total).  For example, over 10$^{14}$ combinations are possible by selecting 25 parameters from a set of 50.  Accordingly, parameter selection is complicated by differing sensitivity, identifiability, and the difficulty of choosing from many combinations.

@@

Local sensitivity analyses showed that \ac{do} was sensitive to perturbations in \Sexpr{nrow(sens_ests_cat)} of 178 parameters (\Sexpr{round(100 * nrow(sens_ests_cat)/178, 0)}\% of total) in FishTank (see \cref{fig:sensalltile} for other state variables). Within each parameter category, \ac{do} was sensitive to \Sexpr{bycat$Optics} parameters for optics (\Sexpr{lncats$Optics} of all optic parameters, \cref{tab:optsens}), \Sexpr{bycat$`Organic Matter`} for organic matter (\Sexpr{lncats$`Organic Matter`}, \cref{tab:omsens}), \Sexpr{bycat$Phytoplankton} for phytoplankton (\Sexpr{lncats$Phytoplankton}, \cref{tab:phytosens}), \Sexpr{bycat$Temperature} for temperature (\Sexpr{lncats$Temperature}, \cref{tab:tempsens}), and \Sexpr{bycat$Zooplankton} for zooplankton (\Sexpr{lncats$Zooplankton}, \cref{tab:zoopsens}). Although \ac{do} had the greatest sensitivity to parameters in the zooplankton category (as percentage of total), the relative effects varied. Among all parameters, sensitivity values ranged from $L1 = $ \Sexpr{sensrng$error[1]} for \Sexpr{sensrng$Parameter[1]} (\Sexpr{tolower(sensrng$Category[1])}) to \Sexpr{sensrng$error[2]} for \Sexpr{sensrng$Parameter[2]} (\Sexpr{tolower(sensrng$Category[2])}), whereas average sensitivity among all parameters was $L1 = $ \Sexpr{sensave}.  Within categories, sensitivity ranged from \Sexpr{sensrngcat$Optics$error[1]} (\Sexpr{sensrngcat$Optics$Parameter[1]}) to \Sexpr{sensrngcat$Optics$error[2]} (\Sexpr{sensrngcat$Optics$Parameter[2]}) for optics, \Sexpr{sensrngcat$`Organic Matter`$error[1]} (\Sexpr{sensrngcat$`Organic Matter`$Parameter[1]}) to \Sexpr{sensrngcat$`Organic Matter`$error[2]} (\Sexpr{sensrngcat$`Organic Matter`$Parameter[2]}) for organic matter, \Sexpr{sensrngcat$Phytoplankton$error[1]} (\Sexpr{sensrngcat$Phytoplankton$Parameter[1]}) to \Sexpr{sensrngcat$Phytoplankton$error[2]} (\Sexpr{sensrngcat$Phytoplankton$Parameter[2]}) for phytopankton, \Sexpr{sensrngcat$Temperature$error[1]} (\Sexpr{sensrngcat$Temperature$Parameter[1]}) to \Sexpr{sensrngcat$Temperature$error[2]} (\Sexpr{sensrngcat$Temperature$Parameter[2]}) for temperature, and \Sexpr{sensrngcat$Zooplankton$error[1]} (\Sexpr{sensrngcat$Zooplankton$Parameter[1]}) to \Sexpr{sensrngcat$Zooplankton$error[2]} (\Sexpr{sensrngcat$Zooplankton$Parameter[2]}) for zooplankton (\cref{fig:sensplo}, bottom).  Average sensitivity values in each category were $L1 = $ \Sexpr{sensave_cat$Optics} for optics, \Sexpr{sensave_cat$`Organic Matter`} for organic matter, \Sexpr{sensave_cat$Phytoplankton} for phytoplankton, \Sexpr{sensave_cat$Temperature} for temperature, and \Sexpr{sensave_cat$Zooplankton} for zooplankton.  Within the six phytoplankton groups, \ac{do} was sensitive to the same parameters within each group although the sensitivity magnitudes varied. Average sensitivity across parameters in each phytoplankton groups showed that \ac{do} was most sensitive to the first and third phytoplankton groups (average $L1 = $ \Sexpr{phytogrps[1, 2]}, \Sexpr{phytogrps[2, 2]}), whereas sensitivity to parameters in the remaining groups was much lower (all with average $L1 < 1$). Sensitivity of \ac{do} did not vary considerably between parameters in the two zooplankton groups (average $L1 = $ \Sexpr{zoopgrps[1, 2]}, \Sexpr{zoopgrps[2, 2]} for groups one and two).
